Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-09 aistats.tex
add table

Exponential Families are to statistics what linear models are to regression.
Many classical model are exponential families, and their convexity and modeling flexibility make them one of the most widely used statistical models. 
Here we review some of their properties. 
We point the reader towards \citet[Chapter 3]{wainwright2008graphical} for a full review.
The exponential family for data $X\in \cX$ with natural (or primal) parameter $\nat$  is specified by a sufficient statistic $T: \cX \rightarrow \real^d$ and a density with respect to a base measure $\nu(dx)$
\begin{equation}
	 p(X|\nat) = \exp( \nat^\top T(X) - \logpart(\nat)) \; ,
\end{equation}
Here $\logpart$ is the log-partition function -- i.e. the normalization term
\begin{align}
    \logpart(\nat) = \log \int e^{\nat^\top T(x)} \nu(dx) \; .
\end{align}
When explicit, we will write the random variable $T = T(X)$ for lightness.
For convenience, we focus on steep exponential families with $\logpart$ a convex function of Legendre type.
In particular, the set $\nat \in \Theta$ such that $\logpart(\nat) < \infty$ is open and convex.
This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}, \nu$ uniform and $T(X)$  the one-hot encoding, and multivariate normal distributions $\cX=\real, \nu$ Lebesgues and $T(X)=(X, X^2)$.

@@

\subsection{Relative Smoothness}
Mirror Descent (MD) \footnote{Mirror Descent is going by many names : Bregman (proximal) Gradient, Relative Gradient Descent, NoLips.} \citep{nemirovski1983problem, beck2003mirror} was first introduced to optimize non-smooth functions, typically under bounded gradient assumption on the objective $f$ or strong convexity assumption on the potential $\logpart$\citep{bubeck}\TODO{bubeck online learning 2011, or monograph?}.

@@

{\bf Traditional results}
in stochastic optimization,
such as the work of \citet{nemirovski2009robust,ghadimi2012optimal},
rely on the assumption that the reference function $A$ 
is itself strongly convex relative to a norm $\norm{\cdot}$,
(meaning that $f$ is strongly convex). 
This apply for the MLE problem of a Gaussian distribution 
with fixed variance, 
but not when the covariance is unknown,
or for a Poisson distribution, as $-\log(\theta)$ and $\exp(\theta)$
are not strongly convex.
Relative smoothness has renewed interest for analyses of mirror descent
without relying on strong-convexity in an underlying norm.
However, they do not yet address the problem satisfactorily.

\TODO{table of results of all 3 papers.}

\textbf{An adaptation of the bounded variance,}
investigated by \citet{hanzely2018fastest}, 
defines the following bounded quantity.
Letting $\theta^+(v)$ be a step from $\theta$ with vector $v$, 
$\theta^+(v) = \nabla A^*(\nabla A(\theta) - v)$,
\alignn{\label{eq:assumption-hanzely}
	\expect[\tilde g]{
	\lin{
		\nabla f(\theta) - \tilde g, 
		\theta^+\!\!\paren{\eta \tilde g } - \theta^+\!\!\paren{\eta \nabla f(\theta)}
	}
	}
	\leq 
	\eta C,
}
for some step-size $\eta$ and constant $C$. Let us unpack this quantity;

{\bf For a Gaussian} with known variance, 
with $A(\theta) = \frac{1}{2}\norm{\theta}^2$, 
this definition recovers the variance of the stochastic gradient
$\expect[\tilde g_t\!\!]{\norm{\nabla f(\theta_t) - \tilde g_t}{}^2}\leq C$.
More generally, if $A$ is $\mu$-strongly convex in $\norm{\cdot}$,
the left-hand side of \cref{eq:assumption-hanzely} is bounded by 
\aligns{
	\frac{1}{\mu} \expect{\norm{\nabla f(\theta_t) - \tilde g_t}{}^2_*},
}
and a uniform bound $C$ exist if the variance of the stochastic gradient is bounded.

{\bf In divergence,} this condition states that the (symmetrized) divergence between 
a deterministic and stochastic update is bounded in expectation,
\aligns{
	&
	\mathbb{E}_{\tilde g_t}\bigg[
	\cB_{A^*}\paren{\mu - \eta g, \mu - \eta \nabla f(\theta)}
	\\
	&\hphantom{\mathbb{E}_{\tilde g}[}
	+ \cB_{A^*}\paren{\mu - \eta \nabla f(\theta), \mu - \eta g}
	\bigg]
	\leq 
	\eta C.
}

{\bf For a maximum likelihood problem, }
where $\tilde g = \mu - x$ 
where $x \sim p(x\cond\theta_*)$,
and $\nabla f(\theta) = \mu - \mu_*$,
\aligns{
	&
	\mathbb{E}_{x}\bigg[
	\cB_{A^*}\paren{(1-\eta) \mu + \eta x, (1-\eta)\mu + \eta \mu_*}
	\\
	&\hphantom{\mathbb{E}_{\tilde g_t}[}
	\cB_{A^*}\paren{(1-\eta)\mu + \eta \mu_*, (1-\eta) \mu + \eta x}
	\bigg]
	\leq 
	\eta C.
}

\fdk{Example where this fails?}

\textbf{Bounded variance at the minimum, }
\fdk{\citep{dragomir2021fast}}

\textbf{Bounded optimality gap,}
\fdk{
Adaptation of the $\expect{\min_x f(x) - \min_x f_i(x)} \leq \sigma^2$ hypothesis \citep{dorazio2021stochastic}}

@@

\fdk{
	Do we use $\alpha, \beta$ anywhere? 
	Might use those for smoothness/strong-convexity rather than $\mu, L$.
} 
\fdk{
	The superscript ${}^*$ denotes the convex conjugate when applied on a function 
	and the optimum when applied to a parameter. 
	How about we use $\theta_n, \theta_*$ for parameter at time $n$ and at the optimum 
	and $f^*$ for the convex conjugate?
}
\fdk{
	On $X_i$ for data; I'm more used to seeing $x_i$ for data, which is the realization of the random variable $X_i$. 
	We don't talk about random variables and don't need the distinction? 
}

>>>>

Exponential Families are to statistics what linear models are to regression.
Many classical model are exponential families, and their convexity and modeling flexibility make them one of the most widely used statistical models. 
Here we review some of their properties. 
We point the reader towards \citet[Chapter 3]{wainwright2008graphical} for a full review.
The exponential family for data $X\in \cX$ with natural (or primal) parameter $\nat$  is specified by a sufficient statistic $T: \cX \rightarrow \real^d$ and a density with respect to a base measure $\nu(dx)$
\begin{equation}
	 p(X|\nat) = \exp( \nat^\top T(X) - \logpart(\nat)) \; ,
\end{equation}
Here $\logpart$ is the log-partition function -- i.e. the normalization term
\begin{align}
    \logpart(\nat) = \log \int e^{\nat^\top T(x)} \nu(dx) \; .
\end{align}
When explicit, we will write the random variable $T = T(X)$ for lightness.
For convenience, we focus on steep, regular exponential families 
where $\logpart$ is strictly convex function of Legendre type,
and the set $\nat \in \Theta$ such that $\logpart(\nat) < \infty$ is open and convex
(See \citet{barndoffnielsen2014information}).
This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}$, $\nu$ uniform and $T(X)$  the one-hot encoding, and multivariate normal distributions $\cX=\real, \nu$ Lebesgues and $T(X)=(X, X^2)$.

@@

Mirror Descent (MD) \footnote{Mirror Descent is going by many names : Bregman (proximal) Gradient, Relative Gradient Descent, NoLips.} \citep{nemirovski1983problem, beck2003mirror} was first introduced to optimize non-smooth functions, typically under bounded gradient assumption on the objective $f$ or strong convexity assumption on the potential $\logpart$\citep{bubeck}\TODO{bubeck online learning 2011, or monograph?}.

\fdk{Online learning;
\citet{azoury2001relative,freund1996predicting,dasgupta2007online}
}

@@

{\bf Traditional results}
in stochastic optimization,
such as the work of \citet{nemirovski2009robust,ghadimi2012optimal},
rely on the assumption that the reference function $A$ 
is itself strongly convex relative to a norm $\norm{\cdot}$,
(meaning that $f$ is strongly convex). 
This apply for the MLE problem of a Gaussian distribution 
with fixed variance, 
but not when the covariance is unknown,
or for a Poisson distribution, as $-\log(\theta)$ and $\exp(\theta)$
are not strongly convex.

Relative smoothness has renewed interest for analyses of mirror descent
without relying on strong-convexity in an underlying norm.
Recent works have made attempts to generalize the assumptions 
that leads to the convergence of mirror descent 
with strongly-convex reference functions, 
such as a uniform bound on the variance of the stochastic gradients $\tilde g(\theta)$,
$\expect*{\norm{\nabla f(\theta) - \tilde g(\theta)}^2} \leq \sigma^2$, 
or only requiring this bounded variance at the minimum $\theta_*$.
However, they do not yet address the problem satisfactorily.
We present those generalization here 
and they are summarized in the context of maximum likelihood estimation 
in \cref{tbl:assumptions}.

\begin{table*}[t]\begingroup
\renewcommand*{\arraystretch}{1.25}%
\centering
\begin{tabular}{p{.35\textwidth}p{.6\textwidth}}
\toprule
Assumption &  \\
\midrule
Strongly convex and bounded variance\newline
\citep[e.g.,][]{??}
&
$
\cB_A(\theta,\theta') \geq \frac{1}{2}\norm{\theta-\theta'}^2,
\quad 
\expect[x]{\norm{\m - x}}^2\leq\sigma^2
$
\\
Uniformly bounded ``variance''\newline
\citep{hanzely2018fastest}
& 
	$
	\expect[x]{
	\begin{aligned}
	\cB_{A^*}\paren{(1-\eta) \m + \eta x, (1-\eta)\m + \eta \m_*}\\
	+\, \cB_{A^*}\paren{(1-\eta)\m + \eta \m_*, (1-\eta) \m + \eta x}
	\end{aligned}
	}
	\leq 
	\eta C
	$
\\
Bounded ``variance'' at $\theta_*$\newline
\citep{dragomir2021fast}
&
	$\expect[x]{
		\cB_{A^*}(\m - 2\eta (\m_* - x), \m)
	} \leq 2 \eta^2 C$
\\
Bounded optimality gap\newline 
\citep{dorazio2021stochastic} 
&
? 
\\
\bottomrule
\end{tabular}
\caption{Caption}
\label{tbl:assumptions}
\endgroup\end{table*}

\subsection{Bounded variance}
An adaptation of the bounded variance,
investigated by \citet{hanzely2018fastest}, 
defines the following bounded quantity.
Letting $\theta^+(v)$ be a step from $\theta$ with vector $v$, 
$\theta^+(v) = \nabla A^*(\nabla A(\theta) - v)$,
\alignn{\label{eq:assumption-hanzely}
	\expect[\tilde g]{
	\lin{
		\nabla f(\theta) - \tilde g(\theta), 
		\theta^+\!\!\paren{\eta \tilde g(\theta)} - \theta^+\!\!\paren{\eta \nabla f(\theta)}
	}
	}
	\leq 
	\eta C,
}
for some step-size $\eta$ and constant $C$. Let us unpack this quantity;

{\bf For a Gaussian} with known variance, 
with $A(\theta) = \frac{1}{2}\norm{\theta}^2$, 
this definition recovers the variance of the stochastic gradient
$\expect[\tilde g_t\!\!]{\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2}\leq C$.
More generally, if $A$ is $\mu$-strongly convex in $\norm{\cdot}$,
the left-hand side of \cref{eq:assumption-hanzely} is bounded by 
\aligns{
	\frac{1}{\mu} \expect{\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2_*},
}
and a uniform bound $C$ exist if the variance of the stochastic gradient is bounded.

{\bf In divergence,} this condition states that the (symmetrized) divergence between 
a deterministic and stochastic update is bounded in expectation,
\alignn{\label{eq:hanzely-in-divergence}
	\begin{aligned}
	&
	\mathbb{E}_{\tilde g}\bigg[
	\cB_{A^*}\paren{\m - \eta \tilde g(\theta), \m - \eta \nabla f(\theta)}
	\\
	&\hphantom{\mathbb{E}_{\tilde g}[}
	+ \cB_{A^*}\paren{\m - \eta \nabla f(\theta), \m - \eta \tilde g(\theta)}
	\bigg]
	\leq 
	\eta C.
	\end{aligned}
}

{\bf For a maximum likelihood problem, }
where $\tilde g(\theta) = \m - x$ 
where $x \sim p(x\cond\theta_*)$,
and $\nabla f(\theta) = \m - \m_*$,
\aligns{
	&
	\mathbb{E}_{x}\bigg[
	\cB_{A^*}\paren{(1-\eta) \m + \eta x, (1-\eta)\m + \eta \m_*}
	\\
	&\hphantom{\mathbb{E}_{\tilde g_t}[}
	\cB_{A^*}\paren{(1-\eta)\m + \eta \m_*, (1-\eta) \m + \eta x}
	\bigg]
	\leq 
	\eta C.
}
\fdk{How do we show that this fails?}

\subsection{Bounded variance at the minimum}

\citep{dragomir2021fast} is similar to \cref{eq:hanzely-in-divergence}, 
where we take only the first part of it 
and care only about gradients at the minimum, replacing 
$\tilde g(\theta)$ and  $\nabla f(\theta)$ with $\tilde g(\theta_*)$ and $\nabla f(\theta_*) = 0$
\aligns{
	\expect[\tilde g]{
		\cB_{A^*}(\m - 2\eta \tilde g(\theta_*), \m)
	} \leq 2 \eta^2 C.
}
{\bf For a maximum likelihood problem, }
\aligns{
	\expect[x]{
		\cB_{A^*}(\m - 2\eta (\m_* - x), \m)
	} \leq 2 \eta^2 C.
}
\fdk{How do we show that this fails?}

\textbf{Bounded optimality gap,}
\fdk{
Adaptation of the $\expect{\min_x f(x) - \min_x f_i(x)} \leq \sigma^2$ hypothesis \citep{dorazio2021stochastic}?
\\
Do they do something with that? It seems to be fixed step-size only.
}

@@

\fdk{
	Do we use $\alpha, \beta$ anywhere? 
	Might use those for smoothness/strong-convexity rather than $\mu, L$.
}

\fdk{
	The superscript ${}^*$ denotes the convex conjugate when applied on a function 
	and the optimum when applied to a parameter. 
	How about we use $\theta_n, \theta_*$ for parameter at time $n$ and at the optimum 
	and $f^*$ for the convex conjugate?
}

\fdk{
	On $X_i$ for data; I'm more used to seeing $x_i$ for data, which is the realization of the random variable $X_i$. 
	We don't talk about random variables and don't need the distinction? 
}

\fdk{
	Need to check 
}
